%\documentclass{acm_proc_article-sp}
\documentclass{sig-alternate}
%\documentclass{vldb}

%%
%% include packages
%%
\usepackage{times}                    %% for time fonts if needed
\usepackage{amsmath}                  %% for math
\usepackage{tabularx}                 %% resize tables
\usepackage{colortbl}                 %% put some color to tables
\usepackage{multirow}                 %% multi-rows/columns on tables
\usepackage{booktabs}                 %% more for tables
\usepackage{tikz}                     %% tikz drawing package
\usepackage{graphicx}                 %% include graphics
\usepackage{algorithm}                %% algorithm floated enviroment
\usepackage{algorithmic}              %% algorithm enviroment
\usepackage{balance}

%%
%% (re)new
%%
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}{Definition}

%%macros
\newcommand{\la}{\langle}
\newcommand{\ra}{\rangle}
\newcommand{\smp}{\mathcal{S}}
\newcommand{\prob}{\mathcal{P}}
\renewcommand{\tt}{\texttt}
\renewcommand{\rm}{\textrm}
\renewcommand{\sf}{\textsf}
\renewcommand{\it}{\textit}
\renewcommand{\bf}{\textbf}
\newcommand{\md}{\textmd}
\renewcommand{\sc}{\textsc}


\begin{document}
\conferenceinfo{SIGMOD'13,} {June 22--27, 2013, New York, New York, USA.}
\CopyrightYear{2013}
\crdata{978-1-4503-2037-5/13/06}
\clubpenalty=10000
\widowpenalty = 10000

\title{Column Imprints: A Secondary Index Structure}
\numberofauthors{2}
\author{
\alignauthor Lefteris Sidirourgos\\
\affaddr{CWI}\\
\affaddr{Amsterdam, The Netherlands}\\
\email{lsidir@cwi.nl}
\alignauthor Martin Kersten\\
\affaddr{CWI}\\
\affaddr{Amsterdam, The Netherlands}\\
\email{mk@cwi.nl}
}

\maketitle

\begin{abstract}
Large scale data warehouses rely heavily on secondary indexes, such as bitmaps
and b-trees, to limit access to slow IO devices. However, with the advent
of large main memory systems, cache conscious secondary indexes are needed to
improve also the transfer bandwidth between memory and cpu. In this paper, we
introduce \it{column imprint}, a simple but efficient cache conscious secondary
index. A column imprint is a collection of many small bit vectors, each
indexing the data points of a single cacheline. An imprint is used during query
evaluation to limit data access and thus minimize memory traffic. The
compression for imprints is cpu friendly and exploits the empirical observation
that data often exhibits local clustering or partial ordering as a side-effect
of the construction process. Most importantly, column imprint compression
remains effective and robust even in the case of unclustered data, while other
state-of-the-art solutions fail. We conducted an extensive experimental
evaluation to assess the applicability and the performance impact of the column
imprints. The storage overhead, when experimenting with real world datasets, is
just a few percent over the size of the columns being indexed. The evaluation
time for over 40000 range queries of varying selectivity revealed the
efficiency of the proposed index compared to zonemaps and bitmaps with WAH
compression.
\end{abstract}

% A category with the (minimum) three required fields
\category{H.3}{Information Storage and Retrieval}{Content Analysis and Indexing}
%\category{H.3}{Information Systems Applications}{Miscellaneous}
%A category including the fourth, optional field follows...
%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]
%\terms{Theory}
\keywords{Secondary Index; Columnar Databases;} % NOT required for Proceedings

\input{introduction}
\input{construction}
\input{operators}
\input{updates}
\input{background}
\input{evaluation}
\input{conclusions}
\bibliographystyle{abbrv}
{\small\bibliography{bibliography}}
%\bibliography{bibliography}
\end{document}
